EBASE manuscript

### 2023-03-15 manu-draft.qmd
maybe better prior comp figure #1

$$
a \sim N(0.2, 0.1)T(0,)
$$ {#eq-aprior}

$$
r \sim N(20, 5)T(0,)
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.01)T(0,0.504)
$$ {#eq-bprior}

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu). Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs. Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c). The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for one day and seven days. Details of these analyses are as follows.

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviation were assessed as changes approximately equal to two orders of magnitude. For example the default value for the mean of the $a$ prior distribution is 0.2 and low and high values were assessed as 0 and 2, respectively. The only exception was the high value that was tested for the mean of the $b$, which was set at 0.504 or twice the default value (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series. For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation. An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period. @fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days. The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE. Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters. Changing the variance of the $a$ parameter had little effect on any of the comparisons. Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter). Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.

The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat. For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D). Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.

@@

\newpage

@@

\newpage

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/apasumdat.png){#fig-apasumdat}

\newpage

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).  The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}

\newpage

@@

\newpage

<!-- \newpage -->

@@

<!-- \newpage -->

>>>>

$$
a \sim N(0, 1)T(0,)
$$ {#eq-aprior}

$$
r \sim N(0, 50)T(0,)
$$ {#eq-rprior}

$$
b \sim N(0, 0.1)T(0,0.502)
$$ {#eq-bprior}

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of water temperature (C) and salinity (psu). Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. An oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created using @eq-ebase as a forward calculating model and user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for one day and seven days to assess how results varied based on the amount of data used for the Bayesian estimation. Details of these analyses are as follows.

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series. The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.504 (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation. An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying somewhat by optimization period. @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days. The results are shown using Root Mean Squared Errors (RMSE) values comparing the synthetic estimates with those from EBASE. Darker colors represent higher RMSE values and can be used to assess how changing the priors affects the ability of EBASE to estimate the true parameter.

Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters. Changing the variance of the $a$ parameter had little effect on any of the comparisons. Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter). Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.

The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-priorcomp. For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D). Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.

@@

{{< pagebreak >}}

@@

{{< pagebreak >}}

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

{{< pagebreak >}}

![Add caption.](figs/priorsumcomp.png){#fig-priorsumcomp}

{{< pagebreak >}}

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-priorcomp).  The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}

{{< pagebreak >}}

@@

{{< pagebreak >}}

<!-- {{< pagebreak >}} -->

@@

<!-- {{< pagebreak >}} -->
